Relative parabolic factorizations of Chevalley groups

### 2015-08-24 bass-kolster.tex
Written proof for Theorem 2

The next theorem is a relative version of the so called Bass---Kolster decomposition (cf.~\cite[Theorem~2.1]{St78}).
\begin{thm}\label{thm:BassKolster}
Let $\Phi$ be a classical root system of rank $\ell\geqslant2$, let $R$ be an arbitrary commutative ring and $I$ be an ideal, satisfying one of the following assumptions:
\[\begin{array}{l@{\quad}l@{\quad}l@{\quad}c}
\Phi = \rA_\ell,\ \ell\geqslant 2, & \sr(I) \leqslant \ell; \\
\Phi = \rC_\ell,\ \ell\geqslant 2, & \sr(I) \leqslant 2\ell-1; \\
\Phi = \rB_\ell, \rD_\ell,\ \ell\geqslant 3, & \asr(I) \leqslant \ell-1.
\end{array}\]
Then the principal congruence subgroup $\G(\Phi,R,I)$ admits the following relative version of Bass---Kolster decomposition:
\[ \G(\Phi,R,I)=  \U(\Phi^+,I) \cdot \U(\Phi^-,I) \cdot Z \cdot \U(\Sigma_1^-\setminus\{-\alpha_\mathrm{max}\},I) \cdot \U(\Sigma_1,I) \cdot \G(\Delta_1,R,I), \]
where $\Z=\left\{ z_{-\alpha_\mathrm{max}}(r,1)\ \middle|\ r\in I \right\}$.
\end{thm}
\begin{proof}

@@

>>>>

The next theorem is a relative version of the so called Bass---Kolster decomposition (cf.~\cite[Theorem~2.1]{St78}).
\begin{thm}\label{thm:BassKolster}
Let $\Phi$ be a classical root system of rank $\ell\geqslant2$, let $R$ be an arbitrary commutative ring and $I$ be an ideal, satisfying one of the following assumptions:
\[\begin{array}{l@{\quad}l@{\quad}l@{\quad}c}
\Phi = \rA_\ell,\ \ell\geqslant 2, & \sr(I) \leqslant \ell; \\
\Phi = \rC_\ell,\ \ell\geqslant 2, & \sr(I) \leqslant 2\ell-1; \\
\Phi = \rB_\ell, \rD_\ell,\ \ell\geqslant 3, & \asr(I) \leqslant \ell-1.
\end{array}\]
Then the principal congruence subgroup $\G(\Phi,R,I)$ admits the following relative version of Bass---Kolster decomposition:
\[ \G(\Phi,R,I)=  \U(\Phi^+,I) \cdot \U(\Phi^-,I) \cdot Z \cdot \U(\Sigma_1^-\setminus\{-\alpha_\mathrm{max}\},I) \cdot \U(\Sigma_1,I) \cdot \G(\Delta_1,R,I), \]
where $Z=\left\{ z_{-\alpha_\mathrm{max}}(r,1)\ \middle|\ r\in I \right\}$.
\end{thm}
\begin{proof}

@@

\begin{proof}[Proof of Theorem~\ref{thm:SL2width}]
Consider the case $\Phi=\rA_\ell$, $\sr(I)\leq 2$. First of all, notice that we can improve the estimate of the number of factors involved in Bass---Kolster decomposition.
Indeed, when performing the first step of the proof of Theorem~\ref{thm:BassKolster} it suffices to make only $2$ additions of $v_{\ell+1}$ (e.\,g. to $v_{1}$ and $v_2$) to make the first $\ell$ entries of $v'$ form a unimodular column.
In particular, $\G(\Phi, R, I)$ can be presented as a product of one element of $\G(\Delta_1, R, I)$, one element of $Z$ and $3\ell+1$ root unipotents $x_\alpha(s)$, $s\in I$.
The latter elements are contained in a product of $3\ell + 1$ copies of $\SL_2(R, I)$. Now the statement of the theorem follows by induction on $\ell$.

The proof in the case $\Phi=\rC_\ell$ is similar (notice that we use the exceptional isomorphism $\SL_2\cong \Sp_2$).
\end{proof}

### 2015-08-24 paper.tex
Written proof for Theorem 2

Another application of parabolic factorizations developed in this paper are subsystem factorizations in terms of subgroups of type $\rA_1$.
The case of a finite field $R=F$ was studied by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{NikProdDecomp}.
More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.
In~\cite{VavKovSL2} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, Gauss decomposition (see~\cite{Sm12}) implies a similar factorization for the elementary subgroup $\E(\Phi, R)$ with $3|\Phi^+|$ factors for arbitrary commutative ring of stable rank $1$.

\begin{thm} Let $I$ be an ideal of a commutative ring $R$ such that $\sr(I)\leq N$. Assume that one of the following is true:
\begin{itemize} 
\item $\Phi=\rA_\ell$, $\ell\geq 2$, $\sr(I)\leq 2$;
\item $\Phi=\rC_\ell$, $\ell\geq 2$, $\sr(I)\leq 3$; \end{itemize}
Then $\G(\Phi, R, I)$ equals the product of at most $3|\Phi^+| + A\rk(\Phi) + B$ copies of subgroups of type $\G(\rA_1, R, I)$. \end{thm}

>>>>

Another application of parabolic factorizations developed in this paper are subsystem factorizations in terms of subgroups of type $\rA_1$.
The case of a finite field $R=F$ was exhaustively studied by M.~Liebeck, N.~Nikolov and A.~Shalev in~\cite{NikProdDecomp}.
More precisely, for an untwisted Chevalley group it has been proved that $\G(\Phi, F)$ is a product of at most $N=5|\Phi^+|$ copies of $\SL(2, F)$.
In~\cite{VavKovSL2} N.~Vavilov and E.~Kovach noted that Bruhat decomposition immediately implies a bound of $N=3|\Phi^+|$ factors for arbitrary field $F$.
In fact, Gauss decomposition (see~\cite{Sm12}) implies a similar factorization for the elementary subgroup $\E(\Phi, R)$ 
with $3|\Phi^+|$ factors for arbitrary commutative ring of stable rank $1$.

In the linear and symplectic case the above results can be extended to even larger class of rings (which includes all Dedekind rings, for example).
\begin{thm}\label{thm:SL2width}
Assume that one of the following is true:
\begin{itemize} 
\item $I$ is an ideal of a (not neccessarily commutative) ring $R$ such that $\sr(I)\leq 2$, $\Phi=\rA_\ell$, $l\geq 2$, $N=3|\Phi^+|+\rk(\Phi) - 3$.
\item $I$ is an ideal of a commutative ring $R$ such that $\sr(I)\leq 3$, $\Phi=\rC_\ell$, $l\geq 2$, $N=3|\Phi^+|+2\rk(\Phi) - 4$
\end{itemize}
Then $\G(\Phi, R, I)$ can be presented as a product of at most $N$ copies of its subgroups isomorphic to $\G(\rA_1, R, I)$. \end{thm}
